Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
varie cose minori

\item \emph{Well Switching rewriting
    systems}\\ Strong Enforcing and existing indepence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

@@

This part is dealt with for well switching rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are well switching
\item Left-linear rewritint systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is well switching
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is well switching
\end{itemize}

@@

We single out a class of left-linear rewriting
systems, referred to as \emph{well-switchable rewriting systems}, 
where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

@@

On the other hand, we show that a concurrency theory of rewriting can
be developed for well-switchable rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Specifically, 
the Church-Rosser theorem is fully recovered.
%for well-switchable rewriting systems.

@@

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switchable systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class of left-linear ones.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.

@@

\subsection{Well-switchable rewriting systems}
\label{subsec:verytame}

@@

\begin{definition}[Well switching rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{well switching} if it is strong enforcing and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence pair between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

The previous definition might sound restrictive. However, it is immediate to see that linear rewriting systems are all well switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are \todo{INSERIRE
  ESEMPI QUI} actually well switching.
%
Roughly speakings, these are possibly hierarchical graphical structures where rules are contrained not to merge elements of top level sorts in the hiearchy (for graphs, nodes can be merged while edges can not).

\begin{definition}
Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph, we define the \emph{depth} of a vertex $v\in V$ in the following way:
\[\dph(v)=\begin{cases}
0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v \} 
\end{cases}\]
\end{definition}

\begin{remark}
Since $\mathcal{G}$ is acyclic and finite, then $\dph(v)$ is a natural number.
\end{remark}

\begin{definition}[Node-merging graphical rewriting systems]\todo{inserire referenze qui}
Let $\mathcal{G}=(V, E, s, t)$ be a directed acyclic graph, the category  $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category $\Set^{\G}$, where $\G$ is the free category over $\mathcal{G}$. A \emph{node-merging graphical rewriting systems} is a left-linear DPO-rewriting system $(\gph{G}, \R)$ such that, for every rule $(l,r)$, with $l\colon K\to L$ and $r\colon K\to R$  in $\R$, the following hold true:
\begin{enumerate}
	\item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$ with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in $\mathcal{G}$) such that  $x$ belongs to the image of $L(p)\colon L(w)\to L(v)$.
	\item $r\colon K\to R$ is mono on vertices of depth $0$: for every $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$ is mono.
\end{enumerate}
\end{definition}

@@

\begin{restatable}{lemma}{lemVTame}
  All node-merging graphical rewriting systems are well switching.
\end{restatable}

@@

\begin{example}[E-graphs]
  Consider the category $\cat{EGraphs}$, where objects are (directed)
  graphs endowed with an equivalence over nodes and arrows are graph
  morphisms which preserve the equivalence, as considered in~\cite{},
  closely related to Egg graphs~\cite{WNW:egg}. Formally,
  $\cat{EGraphs}$ can be seen as the subcategory of the presheaf
  $[E \rightrightarrows V \to Q, \cat{Set}]$ where objects are
  constrained to have the component $Q$ surjective. Explicitly, an
  e-graph $G = \langle s_G, t_G, q_G \rangle$ where
  $s_G, t_G: E_G \rightrightarrows V_G$ provides the graphical
  structure, while the surjective function $q_G : V_G \to Q_G$ maps
  each node to the corresponding equivalence class.

@@

  \todo[inline]{$\cat{EGraph}$ is quasi-adhesive and well switching}  
\end{example}

@@

\begin{example}
  The result above fails when considering left-linear
  derivations. This is readily seen by considering derivations
  $\der{D}$ in Fig.~\ref{fi:derD} and $\der{D}'$ in
  Fig.~\ref{fi:derD1}. We already observed that
  $\der{D} \not\equiv_a \der{D}'$\todo{P: osservato?}
  %
  Still, it is easy to see that they are related by a consistent
  permutation, which is the identity on productions. The colimit of
  both derivations is a single node with a self loop, i.e., the
  terminal object in $\cat{Graph}$, hence all commutation requirements
  are trivially satisfied.
\end{example}

@@

In this section we show, despite the failure of the s, actually much of the theory can
be retained: confluence of switching still holds, while globality of
independence holds in a weak form, conceptually linked to the fact
that fusions introduce a form of disjunctive causality. Finally, a
canonical form for switching sequences can also be recovered.

We start by proving two lemmas dealing with derivations of length
$3$. Despite being technical, these lemma are the fundamental building blocks for showing properties (1) and (2) above.

@@

\begin{example}
  \label{ex:non-global}
  Consider the rewriting system used as running example (rules in
  Fig.~\ref{fi:rules}. Assume to add rules $\rho_3$ and $\rho_4$ as
  depicted in Fig.~\ref{fi:rules-cas}. Consider the derivation
  $\der{F}$ in Fig.~\ref{fi:derF}. Observe that rule $\rho_4$ to be
  applied needs a self loop on node $12$. This can be generated either
  by $\rho_3$ or $\rho_1$ merging nodes $1$ and $2$, hence at least
  one of them must precede the application of $\rho_4$. Indeed, in $\der{E}$ it can be easily seen that the second and third step applying $\rho_1$ and $\rho_3$ are independent. However, we could switch twice the application of $\rho_3$, bringing it in the last position, obtaining a derivation $\der{F}'$ as deipcted in Fig.~\ref{fi:derF1}, where the applications of  $\rho_1$ and $\rho_3$ are no longer independent.

@@

		\item
		\label{lem:indep-global-left:2}
		If, moreover, $(\X, \R)$ is well switching and $S_{i_0, i_1}(\dder{D}_0)\seqind \dder{D}_2$ with an independence pair $(\alpha_0, \alpha_1)$, then
		$S_{i_0,i_1}(\dder{D}_1)\seqind S_{\alpha_0, \alpha_1}(\dder{D}_2)$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}[Confluence of switchings]{lemma}{lemSwitchConfluence}
	\label{lem:switch-confluence}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a
	left-linear DPO rewriting system.
	% 
	Consider a derivation $\der{D}=\{\dder{D}_i\}_{i=0}^2$ and suppose
	that $(i_0,i_1)$ is a strong independence pair between $\dder{D}_0$
	and $\dder{D}_1$, $(a_0,a_1)$ one between $\dder{D}_1$ and
	$\dder{D}_2$ and $(e_0, e_1)$ one between $\dder{D}_0$ and
	$S_{a_0,a_1}(\dder{D}_2)$.\todo{P: non e' meglio qui limitarci a well switching, cosi' da evitare riferimenti a strong pairs?}
\end{restatable}

@@

We showed that the canonical notion of independence adopted for linear
systems does not enjoy some properties that are essential for
developing a sensible theory of concurrency, notably a Church-Rosser
theorem, and we identified a subclass of left-linear rewriting systems,
which we call well-switchable, as an appropriate setting where 
many vital results can be re-established. Specifically, a Church-Rosser 
theorem, i.e. the possibility of performing independent steps in any order, is
fully recovered. Moreover, the switch construction and correspondingly
the switch equivalence, at the core of a theory of concurrency for
rewriting systems, enjoy a number of fundamental properties:
confluence of switching, a weak form of globality of independence and
the existence of a canonical form for switch equivalence proofs.

The class of well-switchable adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we can show that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a class of adhesive categories which include all presheaves over \cat{Set}. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the sorts 
of depth 0  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

The results in this paper open the way to the development of a
concurrent semantics for left-linear rewriting systems over adhesive
categories. In~\cite{baldan2017domains} the authors argue that for
computational systems which allow one to express merging or fusions,
the right semantical models are weak prime domains, a generalisation of
of prime domains, a staple in concurrency theory, and connected event
structures, their event-based counterpart. The mentioned paper
discusses only the case of graph rewriting and does not delve into the
problems induced by the occurring of ``merges'' to the theory of rewriting (the systems
considered there are implicitly assumed to be well-switchable). 
We plan to consolidate the statement that weak prime domains are 
``the'' model for systems with fusions by showing that they allow to provide a semantics
for left-linear rewriting systems in $\mathcal{M}$-adhesive categories.

On the practical side, the present paper just surveyed the possibility
of modelling e-graphs. This appears to be quite interesting as
concurrent rewriting on such structures is an active area of
research. \todo{P: cite} We plan to make the correspondence precise
and investigate how the analysis techniques enabled by a concurrent
semantics of rewriting can impact on e-graphs.

@@

\begin{proposition}
	\label{pr:weak}
	Let $(\X, \R)$ be a linear DPO rewriting system. Then it is well switching.
\end{proposition}
\begin{proof}

@@

		Now, notice that
		\[
		t_1\circ q_1\circ \epsilon'_1  =g_1\circ p_1\circ \epsilon'_1  =g_1\circ a_1 =m_2
		\]
		Hence $(\alpha_0, q_1\circ \epsilon'_1)$ is an independence pair
		for $S_{i_0, i_1}(\dder{D}_0)$ and $\dder{D}_2$. By hypothesis $(\X, \R)$ is well switching and therefore
		$q_1\circ \epsilon'_1$ must coincide with $\alpha_1$.

@@

	By \Cref{def:switch} we also know that $m_0=f'_0\circ a_1$ and $h_1=g'_1\circ a_0$. Now, let us switch $S_{i_0, i_1}(\dder{D}_0)$ and $\dder{D}_2$. In this case we obtain:
	\[\xymatrix@C=15pt{L_1\ar[d]^{f_0\circ i_1}&& K_1 \ar[d]_{k'_1}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){c_0}|(.29)\hole \ar[dr]|(.28)\hole_{h'_1} && L_2 \ar@/_.35cm/[dlll]^(.4){c_1} \ar[dl]|(.28)\hole^{f'_1\circ b_1}& K_2 \ar[d]^{k'_2}\ar@{>->}[l]_{l_2} \ar[r]^{r_2} & R_2 \ar[dr]|(.28)\hole_{h'_2} \ar@/^.35cm/[drrr]_(.4){d_0}|(.29)\hole  && L_0 \ar@/_.35cm/[dlll]^(.4){d_1} \ar[dl]|(.28)\hole^{\hat{m}_0}& K_0 \ar[d]^{\hat{k}_0}\ar@{>->}[l]_{l_0} \ar[rr]^{r_0} && R_0 \ar[d]_{g_2\circ b_0} \\G_0 && \ar@{>->}[ll]^{f'_0} D'_0 \ar[rr]_{g'_0}&& G'_1  && \ar@{>->}[ll]^{\hat{f}_1} \hat{D}_1 \ar[rr]_{\hat{g}_1}&& G'_2 && \ar@{>->}[ll]^{f'_2} D'_2 \ar[rr]_{g'_2}&& G_3 }\]
	Using again the definition of switch, we know that $m'_0=\hat{f}_1\circ d_1$ and $h_2=g'_2\circ d_0$. Moreover, by hypothesis $(\X, \R)$ is well switching and the hypothesis of the second point of \Cref{lem:indep-global-left} are satisfied by $\der{D}$. Thus we can characterize $c_0\colon R_1\to \hat{D}_1$ and $c_1\colon L_2\to D'_0$ as the unique arrows fitting in the diagrams below, where the bottom squares are pullbacks.
	\[\xymatrix@R=15pt@C=15pt{&&R_1 \ar@/_1.2cm/[dddl]_(.4){a_0}|(.63)\hole\ar[dl]^{j_0}\ar[dd]^{c'_0} \ar@{>}[dr]^{c_0}&&&&&L_2 \ar[dl]^{j_1} \ar@/_1.2cm/[dddl]_(.4){z_1}|(.63)\hole\ar[dd]^{c'_1} \ar@{>}[dr]^{c_1}\\&D_2 \ar@{>->}[dl]^(.4){f_2}&&\hat{D}_1 \ar[dr]^{\hat{g}_1}  &&&D_1 \ar@{>->}[dl]^(.4){f_1}&&D'_0 \ar[dr]^{g'_0}\\G_2 && P\ar[dr]_{p_4} \ar@{>->}[ur]_{p_3}\ar@{>->}[dl]^{p_1} \ar[ul]^{p_2}&&G'_2&G_1 && Q\ar[dr]_{q_4} \ar@{>->}[ur]_{q_3}\ar@{>->}[dl]^{q_1} \ar[ul]^{q_2}&&G'_1\\& D'_1 \ar[ul]^{g'_1}&&D'_2 \ar@{>->}[ur]_{f'_2}&&& D_0 \ar[ul]^{g_0}&&D'_1 \ar@{>->}[ur]_{f'_1}}\]

@@

	Moreover, we can characterize the pair $(x_0, x_1)$. If we apply the first point of \Cref{lem:indep-global-left} to $\dder{D}$ we get an independence pair between the last two steps of the derivation above, which, since $(\X, \R)$ is well switching, must coincide with $(x_0, x_1)$. Thus we have two diagrams:

@@

	Finally, switch the second and the third step we get
	\[\xymatrix@C=15pt{L_2\ar[d]^{f_0\circ z_1}&& K_2 \ar[d]_{\mathring{k}_2}\ar@{>->}[ll]_{l_2} \ar[r]^{r_2} & R_2 \ar@/^.35cm/[drrr]_(.4){v_0}|(.29)\hole \ar[dr]|(.28)\hole_{\mathring{h}_2} && L_1 \ar@/_.35cm/[dlll]^(.4){v_1} \ar[dl]|(.28)\hole^{\mathring{f}_1 \circ x_1}& K_1 \ar[d]^{\mathring{k}_1}\ar@{>->}[l]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){u_0}|(.29)\hole\ar[dr]_{\check{h}_1}|(.28)\hole   && L_0 \ar@/_.35cm/[dlll]^(.4){u_1} \ar[dl]^{\mathring{m}_0}|(.28)\hole& K_0 \ar[d]^{\mathring{k}_0}\ar@{>->}[l]_{l_0} \ar[rr]^{r_0} && R_0\ar[d]_{\check{g}_2\circ x_0} \\G_0 && \ar[ll]^{\check{f}_0} \check{D}_0 \ar[rr]_{\check{g}_0}&& \check{G}_1  && \ar@{>->}[ll]^{\bar{f}_1} \bar{D}_1 \ar[rr]_{\bar{g}_1}&& \bar{G}_2 && \ar@{>->}[ll]^{\bar{f}_2} \bar{D}_2 \ar[rr]_{\bar{g}_2}&& G_3 }\]
	As usual, this last switch brings identities $\check{m}_0=\bar{f}_1\circ u_1$ and $g_2\circ j_0=\bar{g}_2\circ u_0$. Applyng the second point of \Cref{lem:indep-global-left} to $\dder{D}_0\cdot S_{j_0, j_1}(\dder{D}_2)\cdot S_{j_0, j_1}(\dder{D}_1)$ and the fact that $(\X, \R)$ is well switching we get the following diagrams:

>>>>

\item \emph{Well-Switching rewriting
    systems}\\ Strong Enforcing and existing indepence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

@@

This part is dealt with for well-switching rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are well-switching
\item Left-linear rewritint systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is well-switching
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is well-switching
\end{itemize}

@@

We single out a class of left-linear rewriting
systems, referred to as \emph{well-switching rewriting systems}, 
where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

@@

On the other hand, we show that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Specifically, 
the Church-Rosser theorem is fully recovered.
%for well-switching rewriting systems.

@@

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class of left-linear ones.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.

@@

\subsection{well-switching rewriting systems}
\label{subsec:verytame}

@@

\begin{definition}[Well-switching rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{well-switching} if it is strong enforcing and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence pair between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

The previous definition might sound restrictive. However, it is immediate to see that linear rewriting systems are all well-switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next show that a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi, modelling of bio and chemical systems) are actually well-switching.\todo{citazioni?}
%
Roughly speakings, these are rewriting systems over possibly
hierarchical graphical structures obtained as algebras of suitable
signatures where rules are constrained not to merge elements of top
level sorts in the hierarchy (for graphs, nodes can be merged while
edges can not). We first introduce the signature which is essentially
an acyclic graph where nodes can be seen as sorts.

\begin{definition}[Graph signature]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph,
  we define the \emph{depth} of a vertex $v\in V$ in the following
  way:
  \[
    \dph(v)=\begin{cases}
      0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
      m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v
      \} 
    \end{cases}
  \]
\end{definition}

Observe that
since $\mathcal{G}$ is acyclic and finite, then $\dph(v)$ is a natural number.

\todo{cite something here? graph signatures?}

\begin{definition}[Node-merging graphical rewriting systems]
  Let $\mathcal{G}=(V, E, s, t)$ be a directed acyclic graph, the
  category $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category
  $\Set^{\G}$, where $\G$ is the free category over $\mathcal{G}$. A
  \emph{node-merging graphical rewriting systems} is a left-linear
  DPO-rewriting system $(\gph{G}, \R)$ such that, for every rule
  $(l,r)$, with $l\colon K\to L$ and $r\colon K\to R$ in $\R$, the
  following hold true:
  \begin{enumerate}
  \item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$
    with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in
    $\mathcal{G}$) such that $x$ belongs to the image of
    $L(p)\colon L(w)\to L(v)$.
  \item $r\colon K\to R$ is mono on vertices of depth $0$: for every
    $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$
    is mono.
  \end{enumerate}
\end{definition}

For instance, in this view, the category $\cat{Graph}$ can be obtained
by taking as $\mathcal{G}$ the graph $e \rightrightarrows v$. The sort
of edges $e$ has depth $0$ while $v$, the sort of nodes, has depth
$1$. Hence condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is mono on edges, i.e., it can only merge nodes.

@@

\begin{restatable}{lemma}{lemVTame}
  All node-merging graphical rewriting systems are well-switching.
\end{restatable}

@@

Another interesting class of well-switching rewriting systems is given by the so-called E-graphs.

\begin{example}[E-graphs]
  Consider the category $\cat{EGraphs}$, where objects are (directed)
  graphs endowed with an equivalence over nodes and arrows are graph
  morphisms which preserve the equivalence, as considered
  in~\cite{BaldanGM06}, closely related to Egg
  graphs~\cite{WNW:egg}. Formally, $\cat{EGraphs}$ can be seen as the
  subcategory of the presheaf
  $[E \rightrightarrows V \to Q, \cat{Set}]$ where objects are
  constrained to have the component $Q$ surjective. Explicitly, an
  e-graph $G = \langle s_G, t_G, q_G \rangle$ where
  $s_G, t_G: E_G \rightrightarrows V_G$ provides the graphical
  structure, while the surjective function $q_G : V_G \to Q_G$ maps
  each node to the corresponding equivalence class.

@@

  \todo[inline]{$\cat{EGraph}$ is quasi-adhesive and well-switching}  
\end{example}

@@

\begin{example}
  The result above fails when considering left-linear
  derivations. This is readily seen by considering derivations
  $\der{D}$ in Fig.~\ref{fi:derD} and $\der{D}'$ in
  Fig.~\ref{fi:derD1}. We already observed that
  $\der{D} \not\equiv_a \der{D}'$\todo{P: osservato?}
  %
  Still, it is easy to see that they are related by a consistent
  permutation, which is the identity on productions. In fact, the colimit of
  both derivations is a single node with a self loop, i.e., the
  terminal object in $\cat{Graph}$, hence all commutation requirements
  are trivially satisfied.
\end{example}

@@

In this section we show that, despite the failure of such
characterisation, actually much of the theory can be retained, taking
a different and sensibly more complex route for the proofs: confluence
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, a canonical form for switching
sequences can also be recovered.

We start by proving two lemmas dealing with derivations of length
$3$. Despite being technical, these lemma are the fundamental building blocks for obtaining our results.

@@

\begin{example}
  \label{ex:non-global}
  Consider the rewriting system used as running example (rules in
  Fig.~\ref{fi:rules}). Assume to add rules $\rho_3$ and $\rho_4$ as
  depicted in Fig.~\ref{fi:rules-cas}. Consider the derivation
  $\der{F}$ in Fig.~\ref{fi:derF}. Observe that rule $\rho_4$ to be
  applied needs a self loop on node $12$. This can be generated either
  by $\rho_3$ or $\rho_1$ merging nodes $1$ and $2$, hence at least
  one of them must precede the application of $\rho_4$. Indeed, in $\der{E}$ it can be easily seen that the second and third step applying $\rho_1$ and $\rho_3$ are independent. However, we could switch twice the application of $\rho_3$, bringing it in the last position, obtaining a derivation $\der{F}'$ as deipcted in Fig.~\ref{fi:derF1}, where the applications of  $\rho_1$ and $\rho_3$ are no longer independent.

@@

		\item
		\label{lem:indep-global-left:2}
		If, moreover, $(\X, \R)$ is well-switching and $S_{i_0, i_1}(\dder{D}_0)\seqind \dder{D}_2$ with an independence pair $(\alpha_0, \alpha_1)$, then
		$S_{i_0,i_1}(\dder{D}_1)\seqind S_{\alpha_0, \alpha_1}(\dder{D}_2)$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}[Confluence of switchings]{lemma}{lemSwitchConfluence}
	\label{lem:switch-confluence}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a
	left-linear DPO rewriting system.
	% 
	Consider a derivation $\der{D}=\{\dder{D}_i\}_{i=0}^2$ and suppose
	that $(i_0,i_1)$ is a strong independence pair between $\dder{D}_0$
	and $\dder{D}_1$, $(a_0,a_1)$ one between $\dder{D}_1$ and
	$\dder{D}_2$ and $(e_0, e_1)$ one between $\dder{D}_0$ and
	$S_{a_0,a_1}(\dder{D}_2)$.\todo{P: non e' meglio qui limitarci a well-switching, cosi' da evitare riferimenti a strong pairs?}
\end{restatable}

@@

We showed that the canonical notion of independence adopted for linear
systems does not enjoy some properties that are essential for
developing a sensible theory of concurrency, notably a Church-Rosser
theorem, and we identified a subclass of left-linear rewriting systems,
which we call well-switching, as an appropriate setting where 
many vital results can be re-established. Specifically, a Church-Rosser 
theorem, i.e. the possibility of performing independent steps in any order, is
fully recovered. Moreover, the switch construction and correspondingly
the switch equivalence, at the core of a theory of concurrency for
rewriting systems, enjoy a number of fundamental properties:
confluence of switching, a weak form of globality of independence and
the existence of a canonical form for switch equivalence proofs.

The class of well-switching adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we showed that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a general class of adhesive categories which include all presheaves over $\cat{Set}$. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the sorts 
of depth 0  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

The results in this paper open the way to the development of a
concurrent semantics for left-linear rewriting systems over adhesive
categories. In~\cite{baldan2017domains} the authors argue that for
computational systems which allow one to express merging or fusions,
the right semantical models are weak prime domains, a generalisation of
of prime domains, a staple in concurrency theory, and connected event
structures, their event-based counterpart. The mentioned paper
discusses only the case of graph rewriting and does not delve into the
problems induced by the occurring of ``merges'' to the theory of rewriting (the systems
considered there are implicitly assumed to be well-switching). 
We plan to consolidate the statement that weak prime domains are 
``the'' model for systems with fusions by showing that they allow to provide a semantics
for left-linear rewriting systems in $\mathcal{M}$-adhesive categories.

On the practical side, the present paper just surveyed the possibility
of modelling e-graphs. This appears to be quite interesting as
concurrent rewriting on such structures is an active area of
research. \todo{P: cite, forse questo? https://arxiv.org/abs/2208.06295} We plan to make the correspondence precise
and investigate how the analysis techniques enabled by a concurrent
semantics of rewriting can impact on e-graphs.

@@

\begin{proposition}
	\label{pr:weak}
	Let $(\X, \R)$ be a linear DPO rewriting system. Then it is well-switching.
\end{proposition}
\begin{proof}

@@

		Now, notice that
		\[
		t_1\circ q_1\circ \epsilon'_1  =g_1\circ p_1\circ \epsilon'_1  =g_1\circ a_1 =m_2
		\]
		Hence $(\alpha_0, q_1\circ \epsilon'_1)$ is an independence pair
		for $S_{i_0, i_1}(\dder{D}_0)$ and $\dder{D}_2$. By hypothesis $(\X, \R)$ is well-switching and therefore
		$q_1\circ \epsilon'_1$ must coincide with $\alpha_1$.

@@

	By \Cref{def:switch} we also know that $m_0=f'_0\circ a_1$ and $h_1=g'_1\circ a_0$. Now, let us switch $S_{i_0, i_1}(\dder{D}_0)$ and $\dder{D}_2$. In this case we obtain:
	\[\xymatrix@C=15pt{L_1\ar[d]^{f_0\circ i_1}&& K_1 \ar[d]_{k'_1}\ar@{>->}[ll]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){c_0}|(.29)\hole \ar[dr]|(.28)\hole_{h'_1} && L_2 \ar@/_.35cm/[dlll]^(.4){c_1} \ar[dl]|(.28)\hole^{f'_1\circ b_1}& K_2 \ar[d]^{k'_2}\ar@{>->}[l]_{l_2} \ar[r]^{r_2} & R_2 \ar[dr]|(.28)\hole_{h'_2} \ar@/^.35cm/[drrr]_(.4){d_0}|(.29)\hole  && L_0 \ar@/_.35cm/[dlll]^(.4){d_1} \ar[dl]|(.28)\hole^{\hat{m}_0}& K_0 \ar[d]^{\hat{k}_0}\ar@{>->}[l]_{l_0} \ar[rr]^{r_0} && R_0 \ar[d]_{g_2\circ b_0} \\G_0 && \ar@{>->}[ll]^{f'_0} D'_0 \ar[rr]_{g'_0}&& G'_1  && \ar@{>->}[ll]^{\hat{f}_1} \hat{D}_1 \ar[rr]_{\hat{g}_1}&& G'_2 && \ar@{>->}[ll]^{f'_2} D'_2 \ar[rr]_{g'_2}&& G_3 }\]
	Using again the definition of switch, we know that $m'_0=\hat{f}_1\circ d_1$ and $h_2=g'_2\circ d_0$. Moreover, by hypothesis $(\X, \R)$ is well-switching and the hypothesis of the second point of \Cref{lem:indep-global-left} are satisfied by $\der{D}$. Thus we can characterize $c_0\colon R_1\to \hat{D}_1$ and $c_1\colon L_2\to D'_0$ as the unique arrows fitting in the diagrams below, where the bottom squares are pullbacks.
	\[\xymatrix@R=15pt@C=15pt{&&R_1 \ar@/_1.2cm/[dddl]_(.4){a_0}|(.63)\hole\ar[dl]^{j_0}\ar[dd]^{c'_0} \ar@{>}[dr]^{c_0}&&&&&L_2 \ar[dl]^{j_1} \ar@/_1.2cm/[dddl]_(.4){z_1}|(.63)\hole\ar[dd]^{c'_1} \ar@{>}[dr]^{c_1}\\&D_2 \ar@{>->}[dl]^(.4){f_2}&&\hat{D}_1 \ar[dr]^{\hat{g}_1}  &&&D_1 \ar@{>->}[dl]^(.4){f_1}&&D'_0 \ar[dr]^{g'_0}\\G_2 && P\ar[dr]_{p_4} \ar@{>->}[ur]_{p_3}\ar@{>->}[dl]^{p_1} \ar[ul]^{p_2}&&G'_2&G_1 && Q\ar[dr]_{q_4} \ar@{>->}[ur]_{q_3}\ar@{>->}[dl]^{q_1} \ar[ul]^{q_2}&&G'_1\\& D'_1 \ar[ul]^{g'_1}&&D'_2 \ar@{>->}[ur]_{f'_2}&&& D_0 \ar[ul]^{g_0}&&D'_1 \ar@{>->}[ur]_{f'_1}}\]

@@

	Moreover, we can characterize the pair $(x_0, x_1)$. If we apply the first point of \Cref{lem:indep-global-left} to $\dder{D}$ we get an independence pair between the last two steps of the derivation above, which, since $(\X, \R)$ is well-switching, must coincide with $(x_0, x_1)$. Thus we have two diagrams:

@@

	Finally, switch the second and the third step we get
	\[\xymatrix@C=15pt{L_2\ar[d]^{f_0\circ z_1}&& K_2 \ar[d]_{\mathring{k}_2}\ar@{>->}[ll]_{l_2} \ar[r]^{r_2} & R_2 \ar@/^.35cm/[drrr]_(.4){v_0}|(.29)\hole \ar[dr]|(.28)\hole_{\mathring{h}_2} && L_1 \ar@/_.35cm/[dlll]^(.4){v_1} \ar[dl]|(.28)\hole^{\mathring{f}_1 \circ x_1}& K_1 \ar[d]^{\mathring{k}_1}\ar@{>->}[l]_{l_1} \ar[r]^{r_1} & R_1 \ar@/^.35cm/[drrr]_(.4){u_0}|(.29)\hole\ar[dr]_{\check{h}_1}|(.28)\hole   && L_0 \ar@/_.35cm/[dlll]^(.4){u_1} \ar[dl]^{\mathring{m}_0}|(.28)\hole& K_0 \ar[d]^{\mathring{k}_0}\ar@{>->}[l]_{l_0} \ar[rr]^{r_0} && R_0\ar[d]_{\check{g}_2\circ x_0} \\G_0 && \ar[ll]^{\check{f}_0} \check{D}_0 \ar[rr]_{\check{g}_0}&& \check{G}_1  && \ar@{>->}[ll]^{\bar{f}_1} \bar{D}_1 \ar[rr]_{\bar{g}_1}&& \bar{G}_2 && \ar@{>->}[ll]^{\bar{f}_2} \bar{D}_2 \ar[rr]_{\bar{g}_2}&& G_3 }\]
	As usual, this last switch brings identities $\check{m}_0=\bar{f}_1\circ u_1$ and $g_2\circ j_0=\bar{g}_2\circ u_0$. Applyng the second point of \Cref{lem:indep-global-left} to $\dder{D}_0\cdot S_{j_0, j_1}(\dder{D}_2)\cdot S_{j_0, j_1}(\dder{D}_1)$ and the fact that $(\X, \R)$ is well-switching we get the following diagrams:
